### 2019-09-17 Thesis/conc.tex
Made some corrections
Random commit.

@@

In this thesis, the base language we have used has been the stateful language Idealized Algol.  
One area to explore is whether the techniques we have developed allow us to build models of effectful versions of stateless languages such as PCF.

One immediate problem we face is that the addition of nondeterministic effects to PCF allows us to distinguish terms that cannot be distinguished within PCF itself.  
Consider, for instance, the PCF term
\[
  M = \lambda x^{\bool}.\If x \Then \Omega \Else (\If x \Then \true \Else \Omega)
  \]
is observationally equivalent (in PCF) to $\Omega$ -- informally, because the term $x$ must have the same `value' both time it is called, so if it is false the first time, then it must be false the second time.
However, the same term inside PCF with finite nondeterminism is not observationally equivalent to $\Omega$: if we substitute nondeterministic choice in for $x$, then it could be false the first time and true the second, causing the term to converge to $\true$.

This means that if $\G$ is a `truly fully abstract' model of PCF -- i.e., one in which observational equivalence of terms corresponds to equality of morphisms rather than intrinsic equivalence\footnote{As an example, take any model that is full abstract in our sense, and take the quotient by the intrinsic equivalence relation.} -- then there can be no natural inclusion functor from $\G$ into a model of PCF with finite nondeterminism.  
Indeed, in such a model $\deno{M}$ and $\deno{\Omega}$ are the same morphism, so they would have to be sent to the same morphism in the model of nondeterministic PCF.

This is a problem for us, since both our techniques -- the Kleisli category and the $\G/\X$ construction -- have a natural functor coming out of the original category $\G$.

All is not lost, however, since we can still try and impose some condition on the original model $\G$ of PCF to say that it is still fine-grained enough to distinguish between certain morphisms that are denotations of observationally equivalent PCF terms.  
One convenient way to do this is to ask that $\G$ should embed into some model $\G^{IA}$ of Idealized Algol.  
An obvious case when this holds is when $\G$ is the (unquotiented) Hyland-Ong category of arenas and innocent strategies, and $\G^{IA}$ is the category of arenas and single-threaded strategies.

For example, fix $X\in\{\bB,\bN\}$ and let \emph{Nondeterministic PCF} be the language PCF, together with a constant $\ask_X \from \bool$ that plays the role of a nondeterministic oracle.  
I.e., it is the language with types given by the following inductive grammar.

@@

  T \Coloneqq \bool \mid \nat \mid T \to T

@@

And with a type theory as shown in Figure \ref{FigNDPcfTypeTheory}.

@@

  However, this does not work if $M$ may exhibit nondeterministic behaviour, since it may evaluate to two different values the two times it is called.

@@

@@

It is a quick check to see that every term of Nondeterministic PCF may be regarded as a term of IA${}_X$, giving us a denotational semantics of the language within $\Kl_{R_X}\G$, where $\G$ is the category of arenas and single-threaded strategies.
But we can be more precise than this: we know from \cite{hoPcf} that the denotation of any term of ordinary PCF is an innocent strategy, so that the denotational semantics of PCF within $\G$ factors through the inclusion $\G_{inn}\hookrightarrow\G$, where $\G_{inn}$ is the category of arenas and innocent strategies.
In particular, the denotation within $\Kl_{R_X}\G$ of any term of ordinary PCF lives within $\Kl_{R_X}\G_{inn}$.
Moreover, the denotation of the new term $\ask_X$, being given by the identity morphism in $\G$, also lives in $\Kl_{R_X}\G_{inn}$, meaning that the whole of Nondeterministic PCF is interpreted within this category.

@@

The passage to full abstraction, via innocent definability for PCF is essentially the same as for Idealized Algol (though if $X=\bN$, then we need to cut down to a category of \emph{recursive} strategies).

@@

At this point, we run into a well-known problem: under what circumstances should a nondeterministic strategy be described as \emph{innocent}? 
The answer that we get from our model is that it is innocent if and only if it takes the form
\[
  A \xrightarrow{\sigma} (X \to B) \xrightarrow{\deno{\ask_X} \to B} B\,,
  \]
where $\sigma$ is an innocent deterministic strategy.  
This is the definition considered -- and ultimately rejected for being too indirect -- by Harmer in his thesis \cite[\sec 3.7]{RusssThesis}.  
This definition is correct, but, perhaps surprisingly, does not coincide with the same definition that we would get by naively applying the usual definition of innocence to nondeterministic strategies.

@@

In the Kleisli category, the innocence of this strategy becomes clear: plays are either of the form
\[
  \begin{tikzcd}[row sep=5pt]
    %
      &
        &
          & q \\
    |[alias=Y]| q  \arrow[urrr, bend left=10] \\
    |[alias=Z]| \true \arrow[u, bend left=45, from=Z.west, to=Y.west] \\
    %
      &
        & q \arrow[uuur, bend left=10]
          & \\
    %
      & |[alias=A]| q \arrow[ur, bend left=10]
        &
          & \\
    %
      & |[alias=B]| \true \arrow[u, bend left=45, from=B.west, to=A.west]
        &
          & \\
    %
      & \vdots
        &
          & \\
    %
      & |[alias=C]| q  \arrow[uuuur, from=C.east, bend right=10]
        & 
          & \\
    %
      & |[alias=D]| \true \arrow[u, bend left=45, from=D.west, to=C.west]
        &
          & \\
    %
      &
        & a \arrow[uuuuuu, bend right=10]
          & \\
    %
      &
        &
          & a \arrow[uuuuuuuuuu, bend right=10]
  \end{tikzcd}
  \]
or of the form
\[
  \begin{tikzcd}[row sep=5pt]
    %
      &
        &
          & q \\
    |[alias=Y]| q  \arrow[urrr, bend left=10] \\
    |[alias=Z]| \false \arrow[u, bend left=45, from=Z.west, to=Y.west] \\
    %
      &
        & q \arrow[uuur, bend left=10]
          & \\
    %
      & |[alias=A]| q \arrow[ur, bend left=10]
        &
          & \\
    %
      & |[alias=B]| \false \arrow[u, bend left=45, from=B.west, to=A.west]
        &
          & \\
    %
      & \vdots
        &
          & \\
    %
      & |[alias=C]| q  \arrow[uuuur, from=C.east, bend right=10]
        & 
          & \\
    %
      & |[alias=D]| \false \arrow[u, bend left=45, from=D.west, to=C.west]
        &
          & \\
    %
      &
        & a \arrow[uuuuuu, bend right=10]
          & \\
    %
      &
        &
          & {a\,,} \arrow[uuuuuuuuuu, bend right=10]
  \end{tikzcd}
  \]
where now the move $\true$ or $\false$ in the nondeterministic oracle game (on the very left) locks the `branching-time information' into the $P$-view.

There is a surprising and elegant definition of nondeterministic innocence (see Levy \cite{levy2014morphisms}, with the correction given by Tsukada and Ong in \cite[Proposition 46]{TsukadaSheaves}), which we shall not detail here.
Tsukada and Ong have also demonstrated that it is instructive to consider strategies as presheaves over plays, where we assign to each individual play a set of possible branches that lead us to that play.  
In this formalism, the innocent nondeterministic strategies are precisely those presheaves that are sheaves with respect to a certain natural Grothendieck topology.

What, then, can our Kleisli category model do for us?
One answer is that it provides a general setting in which we can prove results such as adequacy without worrying about the specific details of a nondeterministic construction.  
For example, we can prove a computational adequacy result for Tsukada and Ong's model $\G_{TO}$ from Corollary \ref{CorMayAdequacyPcfComb} by noting that the denotational semantics in that category factors through the natural functor $\Kl_{R_\bB}\G \to \G_{TO}$.  
Of course, Tsukada and Ong have provided their own proof of Computational Adequacy in the finite nondeterminism case, but this method could be useful if, for example, we wanted to extend their model to deal with countable nondeterminism.

@@

In Chapters \ref{ChapGames} and \ref{ChapFullAbstraction}, we presented a categorical algebra of scoped state via sequoidal categories.  
A natural question to ask is whether we could use the techniques we have developed in the remainder of this thesis in the stateful case.  
Indeed, the state monad
\[
  A \mapsto (W \to (A \times W))
  \]
can be constructed in categories of games.  
More ambitiously, we might hope to capture a parametric notion of local state, via the action
\[
  (R,W).A = R \to (A \times W)\,.
  \]
of $\oppcat \G \times \G$ on $\G$.

@@

One issue, which we have touched on already, is that the state monad is not a monoidal functor, which was the condition we needed in order to ensure that its Kleisli category inherited a monoidal structure.  
Indeed, there is not obvious natural morphism
\[
  (W \to (A \times W)) \times (W \to (B \times W)) \to (W \to ((A \times B) \times W)\,.
  \]

The parametric version suffers from a different problem.  
Suppose we have two \Mellies morphisms
\begin{mathpar}
  f \from A \to (R \to (B \times W))
  \and
  g \from B \to (R' \to (C \times W'))\,.
\end{mathpar}
Then their composition will be given by a \Mellies morphism
\[
  f;g \from A \to ((R \times R') \to (C \times (W \times W')))\,:
  \]
in other words, we will have combined the two inputs $R$ and $R'$ and the two outputs $W$ and $W'$, without any way of specifying that the output state $W$ from $f$ should be used as the input state $R'$ from $g$.

The difficulty is that we might want $W$ and $R'$ to represent different variables, in which case this is the right thing to do, but we might want them to represent the same storage cell, in which case, the output from $W$ should be fed into $R'$.  
It's not clear how to deal with this in a category-theoretic way.

@@

There is also a \emph{local state monad}, due to Plotkin and Power \cite{PlotkinPower} and based on a construction by Levy \cite{PaulsThesis}, which is defined on the category $[\Inj,\Set]$ of functors from $\Inj$, the category of natural numbers and injections, into the category of sets.  
It is given by the following formula, where $V$ denotes a fixes set of values.
\[
  (TA)(n) = V^n \to \left( \int^{p\from\Inj} V^p \times A(p) \times \Inj(n, p) \right)
  \]
Here, if $A \from \Inj \to \Set$ is a functor, then we think of $A(n)$ as the set of all values that $A$ takes on in the presence of $n$ local variables taking values in $V$.

@@

The archetypal Cartesian closed category is the category $\Set$ of sets and functions.  
Most of the work in this thesis has been to do with reader actions (including reader monads as a special case), and we have already provided a classification of the reader actions on $\Set$ -- up to the equivalence relation that identifies actions of monoidal categories $\X$ and $\YY$ if the resulting categories $\Set/\oppcat\X$ and $\Set/\oppcat\YY$ are isomorphic -- in Propositions \ref{PropColimitOfActionIsMonoidalFunctor} and \ref{PropMonoidalFunctorIsColimitOfReaderAction}: namely, they are classified by lax monoidal functors $F\from\Set \to \Set$, in such a way that if a reader action given by an oplax monoidal functor $j\from\X \to \Set$ corresponds to the functor $F \from \Set \to \Set$, then $\Set/\oppcat\X$ is isomorphic to the category $F_*\Set$ formed from $\Set$ by base change along $F$.
Thus, the theory in the case of $\Set$ reduces to the theory of $\Set$-enriched base change.

>>>>

We conclude by examining a few avenues that we are left unexplored in this thesis.

@@

So far, the base language we have used has been the stateful language Idealized Algol.  
One further question to ask is whether the techniques we have developed allow us to build models of effectful versions of stateless languages such as PCF.

One immediate problem we face is that the addition of nondeterministic effects to PCF allows us to distinguish terms that cannot be distinguished within PCF itself.  
For instance, the PCF term
\[
  M = \lambda x^{\bool}.\If x \Then \Omega \Else (\If x \Then \true \Else \Omega)
  \]
is observationally equivalent (in PCF) to $\Omega$ -- informally, because the term $x$ must have the same `value' both time it is called, so if it is false the first time, then it must be false the second time.
However, the same term inside PCF with finite nondeterminism is not observationally equivalent to $\Omega$: if we substitute a nondeterministic oracle in for $x$, then the nondeterministic oracle could return false the first time and true the second, causing the term to converge to $\true$.
Note that this problem does not arise in Idealized Algol: nondeterminism does not allow us to distinguish terms that were indistinguishable in the deterministic language, as we proved in Section \ref{SecIntrinsicEquivalenceRelationKleisli}.

This means that if $\G$ is a `truly Fully Abstract' model of PCF -- i.e., one in which observational equivalence of terms corresponds to equality of morphisms rather than intrinsic equivalence\footnote{As an example, take any model that is full abstract in our sense, and take the quotient by the intrinsic equivalence relation.} -- then there can be no natural inclusion functor from $\G$ into a model of PCF with finite nondeterminism.  
Indeed, in such a model $\deno{M}$ and $\deno{\Omega}$ are the same morphism, so they would have to be sent to the same morphism in the model of nondeterministic PCF.

This is a problem for us, since it means that our model of PCF cannot be the Kleisli category for a monad on $\G$ (or a category of the form $\G/\X$), since then we \emph{do} have a functor coming out of $\G$.

A way to get around this program is to insist that our category $\G$ is not `truly Fully Abstract', but is still fine-grained enough to distinguish between terms that will eventually be distinguished when we add in the nondeterminism.  
A convenient requirement to make is that $\G$ should be embeddable into some model $\G^{IA}$ of Idealized Algol (as is the case for Hyland and Ong's game semantics).  
The idea is that since $\G^{IA}$ must necessarily distinguish between observationally inequivalent terms of nondeterministic PCF, then $\G$ must distinguish between them too.

The other advantage of going via an Idealized Algol model is that we can automatically apply many of the results that we have already obtained.

Let us try this out with an example.  
Fix $X\in\{\bB,\bN\}$ and let \emph{Nondeterministic PCF} be the language PCF, together with a constant $\ask_X \from \bool$ that plays the role of a nondeterministic oracle.  
It is the language with types given by

@@

  T \Coloneqq \bool \mid \nat \mid T \to T\,,

@@

and with a type theory as shown in Figure \ref{FigNDPcfTypeTheory}.

@@

  However, this does not work if $M$ may exhibit nondeterministic behaviour, since $M$ may evaluate to two different values the two times it is called.

@@

This rule is the restriction of the $\converges$ rule from Nondeterministic Idealized Algol to the set of those terms that are terms of Nondeterministic PCF.

@@

It is a quick check to see that every term of Nondeterministic PCF may be regarded as a term of IA${}_X$, giving us a denotational semantics of the Nondeterministic PCF within $\Kl_{R_X}\G$, where $\G$ is the category of arenas and single-threaded strategies.
But we can be more precise than this: we know from \cite{hoPcf} that the denotation of any term of ordinary PCF is an innocent strategy, so that the denotational semantics of PCF within $\G$ factors through the inclusion $\G_{inn}\hookrightarrow\G$, where $\G_{inn}$ is the category of arenas and innocent strategies.
In particular, the denotation within $\Kl_{R_X}\G$ of any term of ordinary PCF lives within $\Kl_{R_X}\G_{inn}$.
Moreover, the denotation of the new term $\ask_X$, being given by the identity morphism in $\G$, also lives in the category
\[
  \G_{inn,X} \coloneqq \Kl_{R_X}\G_{inn}\,,
  \]
giving us a denotational semantics of Nondeterministic PCF within $\G_{inn,X}$.

@@

The passage to full abstraction, via innocent definability for PCF is essentially the same as for Idealized Algol (though, as before, if $X=\bN$, then we need to cut down to a category of \emph{recursive} strategies).

@@

At this point, we run into a well-known problem: under what circumstances should a nondeterministic strategy be described as \emph{innocent}? 
The answer that we get from our model is that it is innocent if and only if it takes the form
\[
  A \xrightarrow{\sigma} (X \to B) \xrightarrow{\deno{\ask_X} \to B} B\,,
  \]
where $\sigma$ is an innocent deterministic strategy.  
This is the definition considered -- and ultimately rejected for being too indirect -- by Harmer in his thesis \cite[\sec 3.7]{RusssThesis}.  
This definition is correct, but, perhaps surprisingly, does not coincide with what we would get by naively applying the usual definition of innocence to nondeterministic strategies.

@@

In the Kleisli category model, the innocence of this strategy becomes clear: plays are either of the form
\[
  \begin{tikzcd}[row sep=5pt]
    %
      &
        &
          & q \\
    |[alias=Y]| q  \arrow[urrr, bend left=10] \\
    |[alias=Z]| \true \arrow[u, bend left=45, from=Z.west, to=Y.west] \\
    %
      &
        & q \arrow[uuur, bend left=10]
          & \\
    %
      & |[alias=A]| q \arrow[ur, bend left=10]
        &
          & \\
    %
      & |[alias=B]| \true \arrow[u, bend left=45, from=B.west, to=A.west]
        &
          & \\
    %
      & \vdots
        &
          & \\
    %
      & |[alias=C]| q  \arrow[uuuur, from=C.east, bend right=10]
        & 
          & \\
    %
      & |[alias=D]| \true \arrow[u, bend left=45, from=D.west, to=C.west]
        &
          & \\
    %
      &
        & a \arrow[uuuuuu, bend right=10]
          & \\
    %
      &
        &
          & a \arrow[uuuuuuuuuu, bend right=10]
  \end{tikzcd}
  \]
or of the form
\[
  \begin{tikzcd}[row sep=5pt]
    %
      &
        &
          & q \\
    |[alias=Y]| q  \arrow[urrr, bend left=10] \\
    |[alias=Z]| \false \arrow[u, bend left=45, from=Z.west, to=Y.west] \\
    %
      &
        & q \arrow[uuur, bend left=10]
          & \\
    %
      & |[alias=A]| q \arrow[ur, bend left=10]
        &
          & \\
    %
      & |[alias=B]| \false \arrow[u, bend left=45, from=B.west, to=A.west]
        &
          & \\
    %
      & \vdots
        &
          & \\
    %
      & |[alias=C]| q  \arrow[uuuur, from=C.east, bend right=10]
        & 
          & \\
    %
      & |[alias=D]| \false \arrow[u, bend left=45, from=D.west, to=C.west]
        &
          & \\
    %
      &
        & a \arrow[uuuuuu, bend right=10]
          & \\
    %
      &
        &
          & {a\,,} \arrow[uuuuuuuuuu, bend right=10]
  \end{tikzcd}
  \]
where now the move $\true$ or $\false$ in the nondeterministic oracle game (on the very left) locks the `branching-time information' into the $P$-view.

Moving away from our Kleisli-category model, there is a surprising and elegant definition of nondeterministic innocence within the usual category of games (see Levy \cite{levy2014morphisms}, with the correction given by Tsukada and Ong in \cite[Proposition 46]{TsukadaSheaves}), which we shall not detail here.
Tsukada and Ong have also demonstrated that it is instructive to consider strategies as presheaves over plays, where we assign to each individual play a set of possible branches that lead us to that play.  
In this formalism, the innocent nondeterministic strategies are precisely those presheaves that are sheaves with respect to a certain natural Grothendieck topology.

Even given this theory, however, our Kleisli category model is still worthwhile, since it provides a general setting in which we can prove results such as adequacy without worrying about the specific details of a nondeterministic construction.  
For example, we can prove a computational adequacy result for Tsukada and Ong's model $\G_{TO}$ from Corollary \ref{CorMayAdequacyPcfComb} by noting that the denotational semantics in that category factors through the natural functor $\Kl_{R_\bB}\G \to \G_{TO}$.  
Tsukada and Ong have provided their own proof of Computational Adequacy in the finite nondeterminism case, but this method could be useful if, for example, we wanted to extend their model to deal with countable nondeterminism.

@@

In Chapters \ref{ChapGames} and \ref{ChapFullAbstraction}, we presented a categorical algebra of scoped state via sequoidal categories.  
The category theory that we used in those chapters is fairly non-standard, and a natural question to ask is whether we can come up with a treatment of state from the point of view of monads or lax actions.

Indeed, there is a well known \emph{state monad}
\[
  A \mapsto (W \to (A \times W))\,,
  \]
which can be constructed in categories of games.  
More ambitiously, we might hope to capture a parametric notion of local state, via the action
\[
  (R,W).A = R \to (A \times W)\,.
  \]
of $\oppcat \G \times \G$ on $\G$.

@@

One issue, which we have touched on already, is that the state monad is not a monoidal functor, which was the condition we needed in order to ensure that its Kleisli category inherited a monoidal structure.  
Indeed, there is no obvious natural morphism
\[
  (W \to (A \times W)) \times (W \to (B \times W)) \to (W \to ((A \times B) \times W))\,.
  \]

The parametric version suffers from a different problem.  
Suppose we have two \Mellies morphisms
\begin{mathpar}
  f \from A \to (R \to (B \times W))
  \and
  g \from B \to (R' \to (C \times W'))\,.
\end{mathpar}
Then their composition will be given by a \Mellies morphism
\[
  f;g \from A \to ((R \times R') \to (C \times (W \times W')))\,:
  \]
in other words, we will have combined the two inputs $R$ and $R'$ and the two outputs $W$ and $W'$.

This works well up to a point, but the main point of state is that we want to be able to write to a variable once and then read from it later.  
So we'd need some way of wiring up the output on the right to the input on the left, and it's not clear how to deal with this in a category-theoretic way.

@@

There is also a \emph{local state monad}, due to Plotkin and Power \cite{PlotkinPower} and based on a construction by Levy \cite{PaulsThesis}, which is defined on the category $[\Inj,\Set]$ of functors from $\Inj$, the category of natural numbers and injections, into the category of sets.  
It is given by the following formula, where $V$ denotes a fixed set of values.
\[
  (TA)(n) = V^n \to \left( \int^{p\from\Inj} V^p \times A(p) \times \Inj(n, p) \right)
  \]
Here, if $A \from \Inj \to \Set$ is a functor, then we think of $A(n)$ as the set of all values that $A$ takes on in the presence of $n$ local variables taking values in $V$.

@@

The archetypal Cartesian closed category is the category $\Set$ of sets and functions.  
Most of the work in this thesis has been to do with reader actions (including reader monads as a special case), and we have already provided a classification of the reader actions on $\Set$ in Propositions \ref{PropColimitOfActionIsMonoidalFunctor} and \ref{PropMonoidalFunctorIsColimitOfReaderAction}: namely, they are classified by lax monoidal functors $F\from\Set \to \Set$, in such a way that if a reader action given by an oplax monoidal functor $j\from\X \to \Set$ corresponds to the functor $F \from \Set \to \Set$, then $\Set/\oppcat\X$ is isomorphic to the category $F_*\Set$ formed from $\Set$ by base change along $F$.
Thus, the theory in the case of $\Set$ reduces to the theory of $\Set$-enriched base change.
